---
title: 'Lab #1: R and Understanding Datasets (30 pts)'
author: "Zully Luna Romero"
format:
  html:
    embed-resources: true
editor: source
knitr:
  opts_chunk:
    message: false
---
```{r}
library(tidyverse)
library(rscorecard)
sc_key("PQF4gDnSgp5jvQz52EfczSEla7urezGtn4eW02yC") # Replace with your own token!!
```
```{r}
scorecard <- sc_init() |>
  sc_year(2022) |>                 #Note how we are looking at only 2024 data here!
  sc_filter(stabbr == "MA") |>     #Note how we are looking at only Massachusetts data here!
 sc_select(unitid, instnm, city, highdeg, control, ugds, adm_rate, costt4_a, costt4_p, pcip27, pctfloan, admcon7, wdraw_orig_yr2_rt, cdr3) |>
sc_get()
```

## Questions

::: {.callout-important}
###### Question 1 (5 pts)
See if you can identify a unique key for this dataset. Write some lines of code to determine whether the column you've identified can act as unique key for the data.

Hint: You need to check whether the values in the column ever repeat.

```{r}
# Write code to calculate number of rows in scorecard
nrow(scorecard)
```


```{r}
# Write code to calculate the number of unique values in the column you've identified as a unique key 
nrow(scorecard)==scorecard$unitid|> #we compared the numbers of rows in the scorecard data set by accessing the column
  unique()|>
  length() 
```

:::

::: {.callout-important}
###### Question 2 (5 pts)
At this point, you've probably figured out that each row in this dataset is a higher education institution. However, there are many different ways that we can define higher education institutions, and that will impact what gets included and excluded in our data. Referencing the Technical Documentation, locate a definition for the unit of observation in this dataset. What institutions are included, and what institutions are excluded? Summarize a definition below.

In the Technical Document, it says that non-Title IV universities are excluded from The Scorecard universe of institutions so their data is not reported. 

:::



::: {.callout-important}

###### Question 3 (5 pts)
For each of the variable names in the `scorecard` data frame, look up the associated name in the data dictionary. You will need to search for the variable name in the sixth column of the data dictionary (I recommend using Ctrl-F to quickly locate the variable name in the spreadsheet.) Once you've found the variable name, reference column 1 to determine what this variable means, and reference columns 7 and 8 to see what possible values will appear in that column.

Identify one nominal variable, one ordinal variable, one discrete variable, and one continuous variable in `scorecard` and list their variable names below. Then uncomment the lines below and use the `typeof()` function to see how `R` determined their data types. Did any surprise you?

```{r}
typeof(scorecard$instnm) #nominal 
typeof(scorecard$highdeg) #ordinal
typeof(scorecard$ugds) #discrete
typeof(scorecard$adm_rate) #continuous
```



:::

::: {.callout-important}
###### Question 4 (5 pts)
Write code below to recode the `admcon7` variable and store the results in a new variable in `scorecard` called `admcon7_text`. You'll need to look up the values in the data dictionary. If you've done this correctly, running this code should produce a barplot that displays multiple bars.

example:
```{r}
scorecard$control_text <-
  recode(
    scorecard$control, 
    "1" = "Public", 
    "2" = "Private nonprofit", 
    "3" = "Private for-profit",
    .default = NA_character_
  )
ggplot(scorecard, aes(x = control_text)) +
  geom_bar()
```

Real code:
```{r}

scorecard$admcon7_text <- 
  recode(
    scorecard$admcon7, 
    "1" = "Required",
    "2" = "Recommended",
    "3" = "Neither required nor recommended",
    "4" = "Do not know",
    "5" = "Considered but not required",
    .default = NA_character_
  )

ggplot(scorecard, aes(x = admcon7_text)) +
  geom_bar()
```

:::

::: {.callout-important}

###### Question 5 (5 pts)
In a code chunk below, calculate the number of missing values in both the `costt4_a` and the `costt4_p` columns. Reference the NOTES column in the data dictionary to determine why there are so many `NA` values in these columns. Add a comment to the code chunk below, explaining the reason for missing values in these columns.

```{r}
sum(is.na(scorecard$costt4_a))
#60
sum(is.na(scorecard$costt4_p))
#115


# According to the NOTES column, a lot are missing because they were discontinued or they exceeded the length or "due to the changing race/ethnicity category definitions
```

:::

::: {.callout-important}

###### Question 6 (5 pts)

Referencing the College Scorecard data documentation, see if you can determine which students are included in calculations of earnings and debt. How might the data’s coverage bias the values that get reported? What might be the social consequences of these biases? Share your ideas on our discussions Slack channel.

The students that are included in the College Scorecard data documentation are the ones that are enrolled in the Title IV universities or students who receive federal grants/loans, according to the Accuracy and Privacy Section of the Technical Document. I think this causes an issue because although they have taken extra measures to insure accuracy for their data, it doesn't not include a variety of people, therefore, limiting the diversity. 
:::

