---
title: "Lab #4: Visualization Aesthetics (30 pts)"
author: "PUT NAME HERE"
format:
  html:
    code-fold: false
execute: 
  echo: true
knitr:
  opts_chunk: 
    message: false
editor: source
---

## Introduction

Use the code below to load and wrangle the data:

```{r}
#| warning: false
#| document: show
library(tidyverse)
library(RColorBrewer)

hmda_ms_2024 <- read.csv("https://ffiec.cfpb.gov/v2/data-browser-api/view/csv?states=MS&years=2024") |>
  filter(loan_type == 1 & 
           lien_status == 1 & 
           conforming_loan_limit == 'C' & 
           derived_dwelling_category %in% c("Single Family (1-4 Units):Site-Built", 
                                            "Single Family (1-4 Units):Manufactured") & 
           business_or_commercial_purpose == 2 &
           loan_purpose %in% c(1, 2)) |>
  select(lei,
         county_code,
         derived_ethnicity,
         derived_race,
         derived_sex,
         action_taken,
         loan_purpose,
         loan_amount,
         interest_rate,
         property_value,
         loan_to_value_ratio,
         denial_reason.1,
         denial_reason.2,
         denial_reason.3,
         denial_reason.4,
         income,
         debt_to_income_ratio) |>
  mutate(loan_purpose = recode(loan_purpose, 
                               "1" = "Home Purchase",
                               "2" = "Home improvement",
                               .default = NA_character_)) |>
  mutate(action_taken = recode(action_taken, 
                               "1" = "Loan originated", 
                               "2" = "Application approved but not accepted",
                               "3" = "Application denied",
                               "4" = "Application withdrawn by applicant",
                               "5" = "File closed for incompleteness",
                               "6" = "Purchased loan",
                               "7" = "Preapproval request denied",
                               "8" = "Preapproval request approved but not accepted",
                               .default = NA_character_)) |>
  mutate(denial_reason.1 = recode(denial_reason.1, 
                                  "1" = "Debt-to-income ratio",
                                  "2" = "Employment history",
                                  "3" = "Credit history",
                                  "4" = "Collateral",
                                  "5" = "Insufficient cash (downpayment, closing costs)",
                                  "6" = "Unverifiable information",
                                  "7" = "Credit application incomplete",
                                  "8" = "Mortgage insurance denied",
                                  "9" = "Other",
                                  "10" = "Not applicable",
                               .default = NA_character_)) |>
  rowwise() |>
  mutate(denial_reason = case_when(any(!is.na(c_across(c(denial_reason.2,
                                                         denial_reason.3,
                                                         denial_reason.4)))) ~ "Multiple",
      TRUE ~ denial_reason.1)) |>
  ungroup() |>
  mutate(loan_to_value_ratio = as.numeric(loan_to_value_ratio),
         property_value = as.numeric(property_value)) |>
  mutate(debt_to_income_ratio = case_when(debt_to_income_ratio %in% c("36", "37", "38","39","40","41") ~ "36%-41%",
                                          debt_to_income_ratio %in% c("42","43","44","45","46","47","48","49") ~ "42%-49%",
                                          TRUE ~ debt_to_income_ratio)) |>
  mutate(debt_to_income_ratio = factor(debt_to_income_ratio, levels = c("Exempt",
                                                                        "<20%",
                                                                        "20%-<30%",
                                                                        "30%-<36%",
                                                                        "36%-41%",
                                                                        "42%-49%",
                                                                        "50%-60%",
                                                                        ">60%"))) |>
  select(-c(denial_reason.1, 
            denial_reason.2, 
            denial_reason.3, 
            denial_reason.4))

```


::: {.callout-important}
###### Question 1

In the code below, create a bar plot that visualizes how many home loans were for home purchases vs. the number for home improvements. Determine the variable that should appear on the x-axis, along with the geom function that can be used to create a bar plot. Fill in the blanks to ensure your plot has the appropriate context.

```{r}
#| eval: false
hmda_ms_2024 |>
  ggplot(aes(x = loan_purpose)) +
  geom_bar() +
  labs(title = "Home Mortgage Disclosure Act Applications, Mississippi, 2024",
       x = "[Home loans for home purchases]",
       y = "[Count of loans]") +
  theme_minimal()
```
:::
- how many loans
- # of home improvements 
::: {.callout-important}
###### Question 2

Create a plot below that counts the number of applicants of each race. Be sure to give it a descriptive title and labels covering all 5 essential components of data context. You can use the code you wrote above as a template to get you started. You'll probably notice that the labels of your plot overlap. You can add `+ coord_flip()` to the end of your code to flip the plot so that the bars are horizontal rather than vertical. This should help with legibility.

```{r}
#| eval: false

# Create plot here
hmda_ms_2024 |>
  ggplot(aes(x = derived_race)) +
  geom_bar() +
  labs(title = "Home Mortgage Disclosure Act Applications, Mississippi, 2024",
       x = "[Race of Homeowners]",
       y = "[# of Applicants]") +
  theme_minimal()+
  coord_flip()




```
- number of applications for race
:::

::: {.callout-important}
###### Question 3

In the code below, create a histogram that visualizes the distribution of loan amounts. Determine the variable that should appear on the x-axis, along with the geom function that can be used to create a histogram. Set the binwidth to create numeric intervals of 100000. Fill in the blanks to ensure your plot has the appropriate context.

```{r}
#| eval: false
hmda_ms_2024  |> 
  ggplot(aes(x = loan_amount)) +
  geom_histogram(binwidth = 100000) +
  labs(title = "Home Mortgage Disclosure Act Applications, Mississippy, 2024",
       x = "[Loan amount]",
       y = "[Count of applicants]",
       caption = str_wrap("This histogram displays the distribution of home loan amounts requested by Mississippi applicants in 2024. The plot shows that the majority of applicants requested loans under $250,000. Since the average home loan size in the U.S. in 2024 was $252,505, this indicates that average requested home loan sizes in MS were smaller in comparison to the country as a whole.", 120)) +
  theme_minimal()
```

- distribution of loan amounts
:::

::: {.callout-important}
###### Question 4

Create a plot below that shows the distribution of loan to value ratios (in intervals of 10). Be sure to give it a descriptive title and labels covering all 5 essential components of data context. Also add a caption to describe the percentage of a loan's property that borrowers most often request when applying for a loan.

You can use the code you wrote above as a template to get you started. Note that there were 20 outlier loans where borrowers requested more than double property's value. There was even one loan where the applicant requested 14 times the property's value. All of these loans were denied. Below I've given you some code to filter out these outliers in order to make the plot more legible. 

```{r}
#| eval: false

hmda_ms_2024 |>
  filter(loan_to_value_ratio < 200) |>
  ggplot(aes(x = loan_to_value_ratio))+
  geom_histogram(binwidth = 10)+
  labs(title = "Home Mortgage Disclosure Act Applications, Mississippy, 2024",
       x = "[Value of loans]",
       y = "[Count of loans]")+
  theme_minimal()

```

- distribution of loans to value ratio
- bin width 10
:::

::: {.callout-important}
###### Question 5

Create a stacked barplot below that counts both the race of the applicants and the action taken in regards to the loan. Be sure to give it a descriptive title and labels covering all 5 essential components of data context. I've given you some code below to move the legend to the bottom of the plot, to reduce the legend size, and to set the number of rows that appear in the legend. 

```{r}
#| eval: false

hmda_ms_2024 |>
  ggplot(aes(x = derived_race, fill = action_taken))+
  geom_bar()+
  labs(title = "Home Mortgage Disclosure Act Applications, Mississipy, 2024",
       x = "Race of Applicants",
       y = "Loan Amount")+
  coord_flip() +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 6), 
        legend.title = element_text(size = 8),
        legend.key.size = unit(0.4, "cm"))+
  
  guides(fill=guide_legend(nrow=4))
  
```

- race of applicants
- action taken in regards of loan
- summarized or individual data
- change aes() to fit 'fill'
:::

::: {.callout-important}
###### Question 6

Copy the plot that you created above into the code chunk below, and then adjust the *position* of the bars to "fill" in order to normalize their height. Add a caption that summarizes your findings. 

```{r}
#| eval: false

hmda_ms_2024 |>
  ggplot(aes(x = derived_race, fill = action_taken))+
  geom_bar()+
  labs(title = "Home Mortgage Disclosure Act Applications, Mississipy, 2024",
       x = "Race of Applicants",
       y = "Loan Amount",
       caption = str_wrap("When the position changes from 'bottom' to 'fill', I noticed it got rid of the legend so the bars weren't as smushed as before", 120))+
  coord_flip() +
  theme(legend.position = "fill",
        legend.text = element_text(size = 6), 
        legend.title = element_text(size = 8),
        legend.key.size = unit(0.4, "cm")) +
  
  guides(fill=guide_legend(nrow=4))

  
```



:::


::: {.callout-important}
###### Question 7
            

Fill in the code below to create a boxplot that visualizes how the distributions of requested loan amounts differed across race. Add a caption that summarizes your findings.  

```{r}
#| eval: false

hmda_ms_2024 |>
  ggplot(aes(x = derived_race, y = loan_amount, fill = derived_race)) +
  geom_boxplot() +
  coord_flip() +
  labs(title = "Home Mortgage Disclosure Act Applications, Mississippi, 2024",
       x = "Race of Applicant",
       y = "Amount of Loans Distributed",
       caption = "I noticed there are many races that are not specifically available which makes their amount of loans distributed, interesting")+
  theme_minimal() 
  
```
- distribution of loan per race 
                         
:::

::: {.callout-important}
###### Question 8
                     CONFUSED
                     
In the code below, fill in the blanks to set the height of each bar to the percent of individuals in a particular race whose mortgage was denied. Select a geom function to create a column chart. Be sure to fill in the correct labels.

```{r}
#| eval: false

hmda_denial_by_race <-
  hmda_ms_2024 |>
  group_by(derived_race) |>
  summarize(total_denied = mean(action_taken == "Application denied"))|>
  ggplot(aes(x = derived_race, y = total_denied)) +
  geom_col() +
  labs(title = "Home Mortgage Disclosure Act Applications, Mississippi, 2024",
       x = "Race of Applicant",
       y = "Application's Denied",
       caption = str_wrap("This column plot visualizes the denial rates of home mortgage applications in MS in 2024. The plot shows that Black or African American applicants, American Indian or Alaska Native applicants, or applicants of 2 or more minority races were all denied mortgage at double the rate of White applicants. This may indicate disparities in lending practices.", 120)) +
  theme_minimal() +
  coord_flip()

hmda_denial_by_race
```


::: {.callout-important}
###### Question 9

Copy your code above, replacing the `hmda_denial_by_race` data frame with `hmda_denial_by_race_sex`. Using the `fill` aesthetic, add the variable for describing the applicants' sex to the plot. Set the position of the columns to `"dodge"`, and be sure to add a label for the `fill` aesthetic. Adjust the caption to summarize your findings.

```{r}
#| eval: false

hmda_denial_by_sex<- 
  hmda_ms_2024 |>
  group_by(derived_sex) |>
  summarize(total_denied = mean(action_taken == "Application denied"))|>
  ggplot(aes(x = derived_sex, y = total_denied, fill = derived_sex)) +
  geom_col(position = "dodge") +
  labs(title = "Home Mortgage Disclosure Act Applications, Mississippi, 2024",
       x = "Sex of Applicant",
       y = "Application's Denied",
       caption = str_wrap("This column plot visualizes the denial rates of home mortgage applications in MS in 2024. The plot shows that a lot of female applicants were denied and very little people who did not make their sex available were denied.", 120)) +
  theme_minimal() +
  coord_flip()

hmda_denial_by_sex  
```

:::

::: {.callout-important}
###### Question 10

Create a plot below that visualizes how many applicants *of each race* were denied mortgages *for each denial reason*. Be sure to give it a descriptive title and labels covering all 5 essential components of data context. You may wish to borrow some of your previous code to adjust the plot for legibility. I've given you some code below to filter the data frame to just those mortgage applications that were denied.

```{r}
#| eval: false

hmda_ms_2024 |>
  filter(action_taken == "Application denied") |>
  group_by(derived_race, denial_reason)|>
  summarize(count_applicants = n())|>
  ggplot(aes(x = denial_reason, y = count_applicants, fill = derived_race))+
  geom_col()+
  labs(
    title ="Denial of Applicants Based off Race",
    x = "Reason for Denail",
    y = "Amount of Applicants Denied",
    fill = "Race of Applicants",
    caption = str_wrap("I noticed there is a big population of Black or African Americans that were denied for either multiple reasons or credit history. The main two races I see are Black or African American and White which is really interesting", 120))+
  theme_minimal()+
  coord_flip()
  
  

  
```
- different applicants based on race
- denied mortgages
- why denied



:::

::: {.callout-important}
###### Question 11

Below I've given you some code to produce a plot that visualizes the percent of mortgage applications that resulted in each action (e.g. loan originated, denied, etc.), broken out by the debt-to-income ratio of the applicant. Referring to last week's lab, add a line to this code to facet this plot by race. For which races was the denial rate higher for applicants with a lower than 50% debt-to-income ratio? Adjust the fact and interpretation presented in the caption with your response.

```{r}
#| eval: true

hmda_ms_2024 |>
  ggplot(aes(x = debt_to_income_ratio, fill = action_taken)) +
  geom_bar(position = "fill") +
  coord_flip() +
  facet_wrap(~derived_race)
  labs(title = "Home Mortgage Disclosure Act Applications, Mississippi, 2024",
       x = "Debt-to-Income Ratio",
       y = "Percent of Mortgage Applications",
       fill = "Action Taken", 
       caption = str_wrap("This bar plot visualizes the percent of mortgage applications that resulted in each action, broken out by the debt-to-income ratio of the applicant. From this plot, we can see that most applicants with a debt-to-income ratio of 50% or higher had a higher denial rate compared to those who had a debt-to-income ratio of 50% or lower.",width = 120)) +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 8), 
        legend.title = element_text(size = 8),
        legend.key.size = unit(0.4, "cm"),
        plot.caption = element_text(size = 8)) +
  guides(fill=guide_legend(nrow = 3))
  
```

:::


::: {.callout-important}
###### Question 12

Below I've given you some code to produce a plot that visualizes how the distributions of interest rates differed across debt-to-income ratios when loans were approved. Referring to last week's lab, add a line to this code to facet this plot by race. For which races were the interest rates higher even with lower debt-to-income ratios?  Adjust the fact and interpretation presented in the caption with your response.


```{r}
#| eval: true

hmda_ms_2024 |>
  filter(action_taken == "Loan originated") |>
  ggplot(aes(x = debt_to_income_ratio, y = as.numeric(interest_rate), fill = derived_race)) +
  geom_boxplot() +
  coord_flip() +
  facet_wrap(~derived_race)
  labs(title = "Home Mortgage Disclosure Act Applications, Mississippi, 2024",
       x = "Debt-to-Income Ratio",
       y = "Interest Rate",
       fill = "Race of Applicants",
       caption = str_wrap("These boxplots visualize the distribution of interest rates for loans originated in MS in 2024, broken out by the debt-to-income ratio of the applicant. From this plot, we can see that Black or African Americans who take out loans tend to have higher interest rates compared to white people and Hawaiin or Pacific Islanders have little to none.",width = 120)) +
  theme_minimal() 
  
```

:::

::: callout-caution
## Ethical Considerations

This dataset makes loan-level data about individual sensitive financial information available to the public. To reduce the risk that borrowers may be identified in the data, the CFPB removes certain variables before publication, such as the loan's unique identifier, the property address, the applicant's credit score, and any free-form text that the applicant writes on the application. Applicants do not have the option to opt-out of their data being disclosed in this dataset. If it were possible to re-identify individual applicants in this dataset, what would be the risks to the public disclosure of their information? Which applicants would be the most vulnerable? How should we think about the balance between the social value of this public data disclosure and the possible risks to those represented in it?  Share your ideas on our `discussions` Slack channel.
:::
